Using best practices for forecasting biodiversity changes in breeding birds

### 2017-07-12 manuscript/draft.Rmd
minor discussion edits

Bird species richness was obtained from the Breeding Bird Survey (BBS)
[@pardieck2017] using the Data Retriever software [@morris2013] and
rdataretriever R package [@mcglinn2017]. The BBS data was filtered to exclude
all nocturnal, cepuscular, and aquatic species [since these species are not well
sampled by BBS methods; @hurlbert2005], as well as unidentified species, and
hybrids. Due to limitations of the environmental data (below), we did not
include any data from outside the continental United States. Observed richness
from 1982 (the first complete year of NDVI data) to 2003 were used for training
the models, and data from 2004 to 2013 was used for testing their performance.
Only BBS routes that were sampled during 70% of the years in the training period
were used (i.e., routes with at least 16 annual observations). All data from
routes that did not meet BBS acceptable criteria were excluded, as were routes
that had not been visited during at least 70% of the years in our model-training
window. The resulting dataset included [[N]] species and [[N]] sites. Among
sites,  richness varied from [[N]] to [[N]] with an average richness of [[N]]
species. Within-site richness variation (among years) was small (standard
deviation of about 5 species, after controlling for observer effects).

Environmental data included a combination of elevation, bioclimatic variables
and a remotely sensed vegetation index (the normalized difference vegetation
index; NDVI), all of which are known to influence richness and distribution in
the BBS data. We used the 4 km resolution PRISM data [@prism] to calculate eight
bioclimatic variables identified as relevant to bird distributions [@harris2015]
for each year in the dataset: mean diurnal range, isothermality, max temperature
of the warmest month, mean temperature of the wettest quarter, mean temperature
of the driest quarter, precipitation seasonality, precipitation of the wettest
quarter, precipitation of the warmest quarter. Satellite-derived NDVI, a primary
correlate of richness in BBS data [@hurlbert2002], with an 8 km resolution was
obtained from the NDIV3g dataset [@pinzon2014] and was available from the time
period 1981-2013. Average summer (May, June, July) and winter (December,
January, Feburary) NDVI values were used as predictors. Elevation was from the
SRTM 90m elevation dataset [@jarvis2008] obtained using the R package raster
[@hijmans2016]. All environmental variables were aggregated to the average
values within a 40 km radius (the length of a BBS route) of the starting point
of each BBS route.

@@

The most commonly used method to forecast future biodiversity, stacked SDMS
[@urban2015], provided the worst overall forecasts of all methods evaluated. In
general, SDM-based forecasts are rarely evaluated, particularly in terms of
repeated observations at the same site. Since site-level richness is relatively
stable, model evaluations that show high accuracy in predicting differences
among sites may not indicate much about a model's ability to predict changes
over time. This issue has also recently been recently highlighted as an issue
for using SDMs to predict the future locations of individual species
[@rapacciuolo2012]. This result is particularly sobering because this approach
forms the main foundation for estimates of the predicted loss of biodiversity to
climate change [@thomas2004; @urban2015]. The poor performance of SDM-based
forecasts highlights the crucial importance of comparing multiple modeling
approaches when conducting ecological forecasts, and in particular the value of
comparing results to simple baselines to avoid over-interpreting the information
present in these forecasts (Box 1).  Disciplines that have more mature forecasting
cultures often do this by reporting "forecast skill", which is the improvement
in the forecast relative to a simple baseline [[weather citation]]. We recommend
following Ye et al. [@ye2015] and adopting this approach in future ecological
forecasting research. It also highlights the need for using hindcasting and
time-series data for assessing the effectiveness of forecasting methods, since
this allows comparisons to baselines and assessment of predicted dynamics within
sites or regions.

@@

Another challenge for ecological forecasting is the timescales of typical
forecasts. We evaluated forecasts using yearly timesteps up to a decade into the
future. In contrast, many ecological forecasts are temporally aggregated to 5-30
year timesteps and projected up to a century into the future. These are commonly
beyond the career or even lifespan of the researchers, making reasonable model
assessment impossible. Currently assessments at these long time-scales can be
only be made with a small number of opportune datasets. For examples hindcasts
have been used to asses models of species richness [@alger2009, distler2015] and
distribution [@rappacciuolo2012, @moran-ordonez2017, @araujo2005,
@eskildsen2013] using data aggregated across similar time scales. More
evaluation studies like these are needed but there is a paucity of adequate long
term data available. Short term forecasts will allow for relatively rapid
validation of models by researchers without needing to wait for enough data to
accumulate [tredennick2016; @dietze2016]. This does not mean that they should
replace large scale long-term forecasts. As discussed above, drivers of species
richness are likely to differ at different temporal scales [@rosenzweig1995,
@white2004, @white2007]. Thus short-term forecasts should be expected to add
value to end users and inform, not replace, long-term forecasts. Hopefully by
evaluating trends in the error shorter term forecasts as done here it will be
possible to also use the information from short-term forecasting to improve
longer-term estimates.

The science of forecasting biodiversity, and ecology more broadly, remains in
its infancy and it is important to consider the general inability the
forecasting methods to improve on simple baselines in that context. When weather
forecasting first started the forecasts were likewise worse than simple
baselines [@mcgill2012]. One of the things that helped weather forecasts improve
was large numbers of forecasts were made in public, which allowed different
approaches to forecasting to be regularly assessed and improved [@mcgill2012;
@silver2012]. This suggests that it is important for ecologists to start
regularly making and evaluating true ecological forecasts, even if they perform
poorly, and to make these forecasts publicly available for assessment. These
forecasts should include both short-term predictions, which can be assessed
quickly, and mid to long-term forecasts to help assess long time-scale processes
and determine how far into the future we can successfully forecast
[@tredenick2016; @dietz2017]. Forecasts based on the models in this paper from
now until 2050 are openly archived on Dryad so that we and others can assess how
well they perform and we plan to evaluate these forecasts and report the results
as each new year of BBS data becomes available. Weather forecasting has
continually improved throughout its history [@bauer2015] in part due to making
and evaluating public forecasts [@mcgill2012] and we hope this will help ecology
do the same.

>>>>

Bird species richness was obtained from the Breeding Bird Survey (BBS)
[@pardieck2017] using the Data Retriever software [@morris2013] and
rdataretriever R package [@mcglinn2017]. The BBS data was filtered to exclude
all nocturnal, cepuscular, and aquatic species
[since these species are not well sampled by BBS methods; @hurlbert2005], as
well as unidentified species, and hybrids. All data from routes that did not
meet BBS acceptable criteria were excluded. Only routes within the continental
United States were used so as to match the climate dataset. The resulting
dataset included [[N]] species and [[N]] sites. Among sites richness varied from
[[N]] to [[N]] with an average richness of [[N]] species. Within-site richness
variation (among years) was small (~5 species).

Environmental data included a combination of elevation, bioclimatic variables
and a remotely sensed vegetation index (the normalized difference vegetation
index; NDVI), all of which are know to influence richness and distribution in
the BBS data. We used the 4 km resolution PRISM data [@prism] to calculate eight
bioclimatic variables identified as relevant to bird distributions [@harris2015]
for each year in the dataset: mean diurnal range, isothermality, max temperature
of the warmest month, mean temperature of the wettest quarter, mean temperature
of the driest quarter, precipitation seasonality, precipitation of the wettest
quarter, precipitation of the warmest quarter. Satellite derived NDVI, a primary
correlated of richness in BBS data [@hurlbert2002], with an 8 km resolution was
obtained from the NDIV3g dataset [@pinzon2014] and was available from the time
period 1981-2013. Average summer (May, June, July) and winter (December,
January, Feburary) NDVI values were used as predictors. Elevation was from the
SRTM 90m elevation dataset [@jarvis2008] obtained using the R package raster
[@hijmans2016]. All environmental variables were aggregated to the average
values within a 40 km radius (the length of a BBS route) of the starting point
of each BBS route.

Data from `r settings$start_yr` (the first complete year of NDVI data) to `r
settings$last_train_year` were used for training models and data from `r
settings$last_train_year + 1` to `r settings$end_yr` was used for testing. Only
BBS routes with at least `r settings$min_num_yrs` annual observations in the
years 1982-2015 were used.

@@

The most commonly used method to forecast future biodiversity, stacked SDMS
[@urban2015], provided the worst overall forecasts of all methods evaluated. In
general, SDM-based forecasts are rarely evaluated, particularly in terms of
repeated observations at the same site. Since site-level richness is relatively
stable, model evaluations that show high accuracy in predicting differences
among sites may not indicate much about a model's ability to predict changes
over time. This issue has also recently been recently highlighted as an issue
for using SDMs to predict the future locations of individual species
[@rapacciuolo2012]. This result is particularly sobering because this approach
forms the main foundation for estimates of the predicted loss of biodiversity to
climate change [@thomas2004; @urban2015]. The poor performance of SDM-based
forecasts highlights the crucial importance of comparing multiple modeling
approaches when conducting ecological forecasts, and in particular the value of
comparing results to simple baselines to avoid over-interpreting the information
present in these forecasts (Box 1).  Disciplines that have more mature forecasting
cultures often do this by reporting "forecast skill", which is the improvement
in the forecast relative to a simple baseline [@jolliffe2003]. We recommend
following Ye et al. [@ye2015] and adopting this approach in future ecological
forecasting research. It also highlights the need for using hindcasting and
time-series data for assessing the effectiveness of forecasting methods, since
this allows comparisons to baselines and assessment of predicted dynamics within
sites or regions.

@@

Another challenge for ecological forecasting is the timescales of typical
forecasts. We evaluated forecasts using yearly timesteps up to a decade into the
future. In contrast, many ecological forecasts are temporally aggregated to 5-30
year timesteps and projected up to a century into the future. These are commonly
beyond the career or even lifespan of the researchers, making reasonable model
assessment impossible. Currently assessments at these long time-scales can
only be made with a small number of opportune datasets. For examples hindcasts
have been used to asses models of species richness [@alger2009, @distler2015]
and distribution [@rappacciuolo2012, @moran-ordonez2017, @araujo2005,
@eskildsen2013] using data aggregated across similar time scales. More
evaluation studies like these are needed but there is a paucity of adequate long
term data available. Short term forecasts will allow for relatively rapid
validation of models by researchers without needing to wait for enough data to
accumulate [@tredennick2016; @dietze2016]. This does not mean that they should
replace large scale long-term forecasts. As discussed above, drivers of species
richness are likely to differ at different temporal scales [@rosenzweig1995,
@white2004, @white2007]. Thus short-term forecasts should be expected to add
value to end users and inform, not replace, long-term forecasts. Hopefully by
evaluating trends in the shorter term forecasts as done here it will be
possible to also use the information from short-term forecasting to improve
longer-term estimates.

The science of forecasting biodiversity, and ecology more broadly, remains in
its infancy and it is important to consider the general inability of
forecasting methods to improve on simple baselines in that context. When weather
forecasting first started the forecasts were likewise worse than simple
baselines [@mcgill2012]. One of the things that helped weather forecasts improve
was large numbers of forecasts were made in public, which allowed different
approaches to forecasting to be regularly assessed and improved [@mcgill2012;
@silver2012]. This suggests that it is important for ecologists to start
regularly making and evaluating true ecological forecasts, even if they perform
poorly, and to make these forecasts publicly available for assessment. These
forecasts should include both short-term predictions, which can be assessed
quickly, and mid to long-term forecasts to help assess long time-scale processes
and determine how far into the future we can successfully forecast
[@tredenick2016; @dietz2017]. Forecasts based on the models in this paper from
now until 2050 are openly archived on Dryad so that we and others can assess how
well they perform and we plan to evaluate these forecasts and report the results
as each new year of BBS data becomes available. Weather forecasting has
continually improved throughout its history [@bauer2015] in part due to making
and evaluating public forecasts [@mcgill2012] and we hope this will help ecology
do the same.
